Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-12-03 _tex/index.tex
Built site for gh-pages

In its most general form, Thurstone's theory deals with pairwise
comparisons of stimuli made by a single judge
\citep[pp.~267]{Thurstone_1927b}. The theory proposes that two key
factors determine the dichotomous outcome of these comparisons: the
discriminal process of each stimulus and their discriminal difference.
The \emph{discriminal process} represents the psychological impact each
stimulus has on judges or, more simply, their underlying perception of
the stimulus' trait level. According to the theory, the discriminal
process for each stimulus follows a Normal distribution, where its mode
(or the mean), referred to as the \emph{modal discriminal process},
indicates the stimulus' position on the trait continuum, and its
dispersion, known as the \emph{discriminal dispersion}, reflects the
variability in the perceived trait level of the stimulus.

@@

Indeed, the theory assumes that the observed dichotomous outcome arises
from the distribution of the difference between the underlying
discriminal processes of the stimuli, known as the \emph{discriminal
difference}. Since the individual discriminal processes follow a Normal
distribution, their difference also follows a Normal distribution
\citep{Andrich_1978}. The mode (or the mean) of this distribution,
representing the (average) relative separation, is given by the
difference between the modal discriminal processes of the stimuli
\(S_{BA} = S_{B} - S_{A}\). Meanwhile, the dispersion of the
distribution, reflecting the variability in the relative separation, is
calculated as
\(\sigma_{BA} = \sqrt{\sigma_{B}^{2} + \sigma_{A}^{2} - \rho\sigma_{B}\sigma_{A}}\).
Here, \(\sigma_{B}\) and \(\sigma_{A}\) are the previously defined
discriminal dispersions, while \(\rho\) measures the correlation between
the discriminal processes of the stimuli. This correlation quantifies
the dependence of the judge's perception of the trait in one stimulus on
his perception of the same trait in the other.

@@

Figure~\ref{fig-correlation} illustrates how varying correlations
influence the distribution of the discriminal difference for the two
texts depicted in Figure~\ref{fig-discriminal_process}. Since the texts
differ in quality, higher correlations increase the likelihood that the
discriminal difference distinctly favors text B over text A. This is
evident from the larger proportion of the area under the curve,
\(P(B > A)\), that lies above zero (shaded gray area in
Figure~\ref{fig-correlation}). Conversely, it is not hard to infer that
if the texts had similar or identical quality levels, higher
correlations would likely reduce the chance that the discriminal
difference distinctly favors one text over the other. This probability
reduction occurs because the distribution of the discriminal difference
would become more narrowly centered around zero (not illustrated).

@@

The previous section outlines the general form of Thurstone's theory,
which applies to a CJ design where a single judge evaluates multiple
stimuli. For the practical application of the theory, Thurstone
developed four additional cases derived from this general form, where
each successive case incorporates additional simplifying assumptions.
Case I represents the general form of the theory. Case II extends this
by allowing multiple judges to make comparisons rather than restricting
them to a single judge. Case III introduces the assumption of zero
correlation between stimuli. Case IV builds on this by assuming that the
stimuli have similar dispersions. Finally, Case V replaces this
assumption with the condition that the stimuli have equal discriminal
dispersions. Table~\ref{tbl-thurstone_cases} summarizes these cases and
their assumptions. For a detailed discussion of this progression, refer
to \citet{Thurstone_1927b} and \citet[pp.~248-253]{Bramley_2008}.

@@

In the first scenario, the intricate aspects of multidimensional,
complex traits may introduce dependencies between the stimuli due to
certain judges' biases that resist cancellation. Research on text
quality suggests that when judges evaluating these traits, they often
rely on various intricate characteristics of the stimuli
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
additional relevant characteristics, which are unlikely to be equally
weighted or opposing, can unevenly influence judges' perceptions,
creating biases in their judgments and, consequently, introducing
dependencies between stimuli
\citep[pp.~346]{vanderLinden_et_al_2017_II}. For example, this could
occur when a judge assessing the argumentative quality of a text places
disproportionate emphasis on grammatical accuracy, ultimately favoring
texts with fewer errors but weaker arguments. While direct evidence for
this specific scenario is lacking, studies such as
\citet{Pollitt_et_al_2003} demonstrate the presence of such biases,
supporting the idea that the factors influencing pairwise comparisons
may not always cancel out.

@@

In any case, the psychometric and statistical literature underscores the
importance of incorporating additional relevant traits and clustering
(grouping) structures into the modeling of CJ data to address potential
dependencies between stimuli. Oversights in these areas can lead to
misleading conclusions about the trait's reliability and ultimately
compromise its validity \citep[pp.~2]{Perron_et_al_2015}. Neglecting
additional traits relevant to the stimuli or the trait of interest, such
as judges' biases, introduces a dimensional mismatch in the modeling
process of the CJ data. This mismatch often results in an overestimation
of the trait's reliability \citep[pp.~341]{Hoyle_et_al_2023} or, worse,
introduces biases into the estimated trait \citep{Ackerman_1989}.
Therefore, these traits must be incorporated into the modeling of CJ
data, as demonstrated by \citet{Andrich_1978} and
\citet{Wainer_et_al_1978}. Similarly, ignoring hierarchical (grouping)
structures reduces the accuracy of model parameter estimates and further
exacerbates the overestimation of reliability
\citep[pp.~482]{Hoyle_et_al_2023}. To address this, any additional
hierarchical structure relevant to the stimuli should also be included
in the model, as in Multilevel Structural Equation models (MSEM)
\citep[chap.~26]{Hoyle_et_al_2023}.

>>>>

In its most general form, Thurstone's theory deals with pairwise
comparisons of stimuli made by a single judge
\citep[pp.~267]{Thurstone_1927b}. The theory proposes that two key
factors determine the dichotomous outcome of these comparisons: the
discriminal process of each stimulus and their discriminal difference.
The \emph{discriminal process} represents the psychological impact each
stimulus has on judges or, more simply, their underlying perception of
the stimulus' trait level. According to the theory, the discriminal
process for each stimulus follows a Normal distribution, where its mode
(mean), referred to as the \emph{modal discriminal process}, indicates
the stimulus' position on the trait continuum, and its dispersion, known
as the \emph{discriminal dispersion}, reflects the variability in the
perceived trait level of the stimulus.

@@

Indeed, the theory assumes that the observed dichotomous outcome arises
from the distribution of the difference between the underlying
discriminal processes of the stimuli, known as the \emph{discriminal
difference}. Since the individual discriminal processes follow a Normal
distribution, their difference also follows a Normal distribution
\citep{Andrich_1978}. The mode (mean) of this distribution, representing
the (average) relative separation, is given by the difference between
the modal discriminal processes of the stimuli
\(S_{BA} = S_{B} - S_{A}\). Meanwhile, the dispersion of the
distribution, reflecting the variability in the relative separation, is
calculated as
\(\sigma_{BA} = \sqrt{\sigma_{B}^{2} + \sigma_{A}^{2} - \rho\sigma_{B}\sigma_{A}}\).
Here, \(\sigma_{B}\) and \(\sigma_{A}\) are the previously defined
discriminal dispersions, while \(\rho\) measures the correlation between
the discriminal processes of the stimuli. This correlation quantifies
the dependence of the judge's perception of the trait in one stimulus on
his perception of the same trait in another.

@@

Figure~\ref{fig-correlation} illustrates how varying correlations
influence the distribution of the discriminal difference for the two
texts depicted in Figure~\ref{fig-discriminal_process}. Since the texts
differ in quality, higher correlations increase the likelihood that the
discriminal difference distinctly favors text B over text A. This is
evident from the larger proportion of area under the curve,
\(P(B > A)\), that lies above zero (shaded gray area in
Figure~\ref{fig-correlation}). Conversely, it is not hard to infer that
if the texts had similar or identical quality levels, higher
correlations would likely reduce the chance that the discriminal
difference distinctly favors one text over the other. This probability
reduction occurs because the distribution of the discriminal difference
would become more narrowly centered around zero (not illustrated).

@@

The previous section outlines the general form of Thurstone's theory,
which applies to a CJ design where a single judge evaluates multiple
stimuli. For the practical application of the theory, Thurstone
developed four additional cases derived from this general form, where
each successive case incorporates additional simplifying assumptions.
Case I represents the general form of the theory. Case II extends this
by allowing multiple judges to make comparisons rather than restricting
the comparisons to a single judge. Case III introduces the assumption of
zero correlation between stimuli. Case IV builds on this by assuming
that the stimuli have similar dispersions. Finally, Case V replaces this
assumption with the condition that the stimuli have equal discriminal
dispersions. Table~\ref{tbl-thurstone_cases} summarizes these cases and
their assumptions. For a detailed discussion of this progression, refer
to \citet{Thurstone_1927b} and \citet[pp.~248-253]{Bramley_2008}.

@@

In the first scenario, the intricate aspects of multidimensional,
complex traits may introduce dependencies between the stimuli due to
certain judges' biases that resist cancellation. Research on text
quality suggests that when judges evaluate these traits, they often rely
on various intricate characteristics of the stimuli to form their
judgments
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
additional relevant characteristics, which are unlikely to be equally
weighted or opposing, can unevenly influence judges' perceptions,
creating biases in their judgments and, consequently, introducing
dependencies between stimuli
\citep[pp.~346]{vanderLinden_et_al_2017_II}. For example, this could
occur when a judge assessing the argumentative quality of a text places
disproportionate emphasis on grammatical accuracy, ultimately favoring
texts with fewer errors but weaker arguments. While direct evidence for
this specific scenario is lacking, studies such as
\citet{Pollitt_et_al_2003} demonstrate the presence of such biases,
supporting the idea that the factors influencing pairwise comparisons
may not always cancel out.

@@

In any case, the psychometric and statistical literature emphasizes the
importance of addressing factors that create dependencies between
stimuli, as failing to do so can lead to inaccurate conclusions about a
trait's reliability and, by extension, its validity. For instance,
neglecting additional traits relevant to the stimuli, such as judges'
biases, often leads to a dimensional mismatch in the model. This
mismatch can potentially inflate the reliability of the trait
\citep[pp.~341]{Hoyle_et_al_2023} or, worse, introduce bias into the
trait's estimates \citep{Ackerman_1989}. Similarly, failing to account
for hierarchical (grouping) structures reduces the precision of model
parameter estimates, further amplifying the overestimation of
reliability \citep[pp.~482]{Hoyle_et_al_2023}. These issues collectively
undermine the trait's reliability, and because ``validity cannot exist
without reliability'' \citep[pp.~2]{Perron_et_al_2015}, they ultimately
compromise the validity of results derived from the the data modeling.

Fortunately, the same literature provides solutions for addressing the
factors that create dependencies between stimuli. \citet{Andrich_1978}
and \citet{Wainer_et_al_1978} propose integrating judges' biases into
the CJ model. This integration would produce a model similar to a
multidimensional or two-parameter logistic IRT model
\citep[chap.~15]{Hoyle_et_al_2023}, depending on the theoretical
assumptions a researcher is willing to make regarding the judges'
biases. The literature also suggests incorporating relevant hierarchical
structures into the model to account for dependencies in the stimuli,
resulting in a model resembling a Multilevel Structural Equation Model
(MSEM) \citep[chap.~26]{Hoyle_et_al_2023}.
